---
title: "Visualizing Premier League fixture difficulty"
author: "Didrik Finnøy"
date: 2019-08-14
tags: ["R"]
series: ["Fantasy Premier League"]
series_weight: 2
description: "The tone within one of my fantasy leagues has become rather sassy as of late."
images: ["didjfin-cursive.png"]
---

When playing Fantasy Premier League, people are very much concerned with what opposition the players on their team will be facing.
A commonly held belief is that players tend to get more fantasy points against weaker opposition.
I haven't done any actual analysis to support this belief, but it sounds perfectly reasonable, and I think it's wise to keep upcoming fixtures in mind when planning your transfers.

With that in mind, wouldn't it be neat if it was super easy to quickly get an overview over every single team's schedule?
If you don't subscribe to any premium FPL analysis services, you're probably evaluating matchups and fixture difficulties by clicking through the FPL user interface.
The purpose of this post is to show you that there's a better way, you'll also find a nifty data visualization that you can use for the remainder of the 2019-20 Premier League season.

## Getting data on upcoming fixtures
Let's get some data from the Premier League API.
If you're not familiar with this data source, check out the first post in this series (there is a link at the top of this page).
```{r, eval = F}
library(tidyverse)
library(jsonlite)

# Download the `bootstrap-static` dataset into your R session
url <- "https://fantasy.premierleague.com/api/"
bootstrap_static <- fromJSON(paste0(url, "bootstrap-static"))
```
```{r, echo = F, message = F}
library(tidyverse)
library(jsonlite)
bootstrap_static <- readRDS("./data/bootstrap-static-120819.RDS")
```
I'm aiming to keep this post self-contained, and not rely on any code or data produced in previous blog posts.
The `bootstrap_static` dataset from the FPL API contains all the information we need for superior fixture planning that will shock and baffle your competition.
We'll start by creating an overview of the teams participating in the Premier League this season
```{r}
teams <- bootstrap_static$teams %>% 
  as_tibble() %>% 
  select(name, id, code, starts_with("strength"))

teams
```
Lots of juicy information here. 
Not only does this dataset allow us to evaluate players based on the opponents overall strength; we can also compare offensive versus defensive strength, and home versus away effects.
We don't even have to calculate anything, all that needs to be done is present preexisting information in a user friendly manner.

The next piece of the puzzle is to get an overview of the Premier League schedule, this data is also contained within the FPL API, but not within the `bootstrap_static` object.
To get a hold of this information, we need to access the extended player data available in the `element-summary` branch of the API. 
We only need one of these summaries from each team, so the next step will be to pick a player from each team, and download his summary data.

```{r, eval = F}
# Get the `id` numer of one player from each team
team_fixtures <- bootstrap_static$elements %>% 
  as_tibble() %>% 
  group_by(team_code) %>% 
  summarize(player_id = first(id)) %>% 
  ungroup() %>% 
  # Add team names
  left_join(
    teams %>% rename(team = name, team_code = code), by = "team_code"
  ) %>% 
  # Download an `element-summary` for each `player_id`
  mutate(
    element_summary = map(player_id, function(x) {
      paste0(
        "https://fantasy.premierleague.com/api/element-summary/", x, "/"
      ) %>% 
        fromJSON(.)
    }),
    # Extract the fixture overview from each `element_summary`
    fixtures = map(element_summary, function(x) x$fixtures)
  ) %>% 
  # Drop columns that we don't need anymore
  select(-team_code, -id, -player_id, -element_summary)

team_fixtures
```
```{r, echo = F}
team_fixtures <- readRDS("./data/team_fixtures-140819.RDS")
team_fixtures
```

At the moment, each team's fixtures are stored within nested datasets under the `fixtures` column.
The attentive reader will notice that each team only has 37 fixtures contained within their `fixtures` dataset, not 38 which is the number of games they play over the course of a season.
Why? Because the particular dataset we've used from the API does not include completed fixtures, and at the time of writing, one round of the Premier League has already been played.

Our next step will be to clean up the nested `fixtures` datasets under `team_fixtures`:
```{r}
# Display one of our 20 `fixtures` datasets
team_fixtures$fixtures[[1]] %>% as_tibble()
```
The nested datasets contain a lot of information that we're not interested in right now, and the data's format is not optimal for creating nice visualizations.
A little bit of data wrangling is required before we proceed:

```{r}
plot_data <- team_fixtures %>% 
  mutate(
    # Transform the nested `fixtures` datasets
    fixtures = map(fixtures, function(x) {
      x %>%
        # Get opponent's id number
        mutate(opponent_team_id = ifelse(is_home, team_a, team_h)) %>% 
        # Only keep the variables we need
        select(
          gameweek = event, opponent_team_id, is_home, difficulty
        ) %>% 
        # Merge with the `teams` dataset for team names and strength variables
        left_join(
          teams %>% 
            # Pick variables, and rename them since they're reflecting
            # the opposition
            select(
              opponent_team_id = id, 
              opponent_team = name,
              opponent_strength = strength,
              opponent_strength_overall_home = strength_overall_home,
              opponent_strength_overall_away = strength_overall_away,
              opponent_strength_attack_home = strength_attack_home,
              opponent_strength_attack_away = strength_attack_away,
              opponent_strength_defence_home = strength_defence_home,
              opponent_strength_defence_away = strength_defence_away
            ),
          by = "opponent_team_id"
        ) %>% 
        select(-opponent_team_id)  # Don't need this anymore
    })
  ) %>% 
  # Convert from nested to long format, leaving us with one row per team-fixture
  unnest() %>% 
  # Impose more practical column ordering
  select(
    gameweek, team, opponent_team, is_home, difficulty, everything()
  )

plot_data %>% as_tibble()
```
That's more like it, now we have a dataset, `plot_data` thats easily digestable by our plotting packages.

## Visualizing the data

My preferred way of visualizing this information is through a *heatmap*, also known as a *tile plot*.
I figured it would be useful to demonstrate three techniques for producing this type of plot with R.
We'll start with the simplest approach, and finish with the most complex one.

##### Easy heatmap with `ggplot2`

When I just want a static plot with no bells and whistles, `ggplot2` is my preferred visualization package.
The package has already been loaded through the `library(tidyverse)` command earlier on.
You can of course load it explicitly, without the other `tidyverse` packages, by running the `library(ggplot2)` command in your R session.
I'll be using the `RColorBrewer` package for a nice-looking and intuitive color scheme, and the `ggthemes` package to make the plot look nicer without having to write a lot of code.
```{r, out.width = "100%"}
library(RColorBrewer)
library(ggthemes)

plot <- plot_data %>% 
  # Modify the `team` variable so teams are displayed in alphabetical order
  mutate(
    team = factor(team) %>%
      factor(., levels = rev(levels(.)))
  ) %>% 
  # Create the plot based on our input data
  ggplot(aes(x = gameweek, y = team, fill = difficulty)) + 
  geom_raster() +
  # Specify the x-axis layout
  scale_x_continuous(
    limits = c(0, 39), 
    breaks = seq(2, 38, 2), 
    expand = c(0, 0)
  ) +
  # Set the color scheme
  scale_fill_distiller(palette = "YlOrRd", direction = 1) +
  # Set the theme
  theme_tufte(base_family = "sans-serif") +
  # Set the labels
  labs(
    title = "Fixture difficulty, Premier League 2019-20",
    x = "Gameweek", 
    y = "", 
    fill = "Difficulty"
  )

plot
```
Not too shabby, but we can do better.


##### Interactive heatmap with `plotly`

I want to to incorporate even more information into the plot without making it too noisy.
We could, for example, make it so that hovering your cursor over one of the colored tiles would provide information about that fixture.
This is quite easy to do with the excellent `plotly` package for R.
The automatic sizing feature of the `ggplotly()` function also makes the plot look a whole lot nicer:

```{r, message = F, warning = F, out.width = "100%"}
library(plotly)

plot <- plot_data %>% 
  # Modify the `team` variable so teams are displayed in alphabetical order
  mutate(
    team = factor(team) %>%
      factor(., levels = rev(levels(.)))
  ) %>% 
  # Create the plot based on our input data
  ggplot(aes(x = gameweek, y = team, fill = difficulty)) + 
  geom_raster(
    # Define the tooltip contents with HTML
    aes(text = paste0(
      "<b>", team, " (GW", gameweek, 
      ifelse(is_home, " Home", " Away"), ")</b><br>",
      "Attack strenght<br>",
      "&nbsp;&nbsp;Home: ", strength_attack_home, "<br>",
      "&nbsp;&nbsp;Away: ", strength_attack_away, "<br>",     
      "Defense strength<br>",
      "&nbsp;&nbsp;Home: ", strength_defence_home, "<br>",
      "&nbsp;&nbsp;Away: ", strength_defence_away, "<br><br>",     
      "<b>vs. ", opponent_team, "</b><br>",
      "Attack strenght<br>",
      "&nbsp;&nbsp;Home: ", opponent_strength_attack_home, "<br>",
      "&nbsp;&nbsp;Away: ", opponent_strength_attack_away, "<br>",     
      "Defense strength<br>",
      "&nbsp;&nbsp;Home: ", opponent_strength_defence_home, "<br>",
      "&nbsp;&nbsp;Away: ", opponent_strength_defence_away
    ))
  ) +
  # Specify the x-axis layout
  scale_x_continuous(
    limits = c(0, 39), 
    breaks = seq(2, 38, 2), 
    expand = c(0, 0)
  ) +
  # Set the color scheme
  scale_fill_distiller(palette = "YlOrRd", direction = 1) +
  # Set the theme
  theme_tufte(base_family = "sans-serif") +
  # Set the labels
  labs(
    title = "Fixture difficulty, Premier League 2019-20",
    x = "Gameweek", 
    y = "", 
    fill = "Difficulty"
  )

# Render the `ggplot2` visualization with `plotly`
ggplotly(plot, tooltip = "text") %>% 
  # Disable unneccessary plotly functionality
  layout(
    xaxis = list(fixedrange = T),  
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F) 
```

Useful, but we can do even better.
When we're looking at upcoming fixtures, trying to figure out what players we should add to our fantasy teams, we might only be interested one team's attacking strength, versus another team's defensive strength.
The plot above allows you to compare these variables, but it requires too many cognitive calories for my liking.


##### Even more interactive heatmap with `plotly` + `shiny`

We can take this plot to the next level by allowing users to choose what information they would like to see.
The go-to library for such matters in the world of R, is called `shiny`.
Making a `shiny` application requires a fair bit of code; too much for this post.
Instead, I'll let you play around with the end result below, and refer you to my GitHub repository if you'd like to see how it was made.
You'll find an R notebook (.rmd) file that you can copy-paste into RStudio, and run to create an up-to-date version of the plot below.
It should totally work for the remainder of the 2019-20 season (assuming you've installed all the required packages).
If they change the API again next year, it might need some adjustments.


<iframe src="http://www.didjfin.no/shiny/fpl/visualizing-fixture-difficulty/" scrolling="no" frameborder="no" style='height:1750px;'></iframe>